Insecticide resistance evolution with mixtures and sequences : a model-based introduction.

### 2017-08-11 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.tex
paper2 writing

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Insecticide resistance evolution with mixtures and sequences : a model-based graphical introduction and summary},
            pdfauthor={Andy South and Ian Hastings},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Insecticide resistance evolution with mixtures and sequences : a
model-based graphical introduction and summary}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Andy South and Ian Hastings}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{2017-04-11}

@@

\subsubsection{Insecticide effectiveness predicted to be most important
factor determining whether insecticide mixtures can slow evolution of
resistance.}\label{insecticide-effectiveness-predicted-to-be-most-important-factor-determining-whether-insecticide-mixtures-can-slow-evolution-of-resistance.}

Target : Malaria journal (other suggestions welcome)

@@

\emph{must be 350 words, currently 351}

@@

Insecticide resistance threatens the control of vectors and particularly
malaria. New insecticides are being developed to address this. Potential
strategies for using new insecticides include mixtures, sequences and
rotations and there are recommendations for how best to implement these
to manage resistance. There is, however, limited recent, accessible
modelling work addressing the evolution of resistance under different
operational strategies. Previous work concludes that preferred
strategies will be situation specific. There is potential to improve the
level of mechanistic understanding within the operational community of
how insecticide resistance can be expected to evolve in response to
different strategies.

This paper develops an accessible, mechanistic picture of how the
evolution of insecticide resistance is likely to respond to potential
intervention strategies to help guide both management and policy. The
aim is to reach an audience unlikely to read a more detailed modelling
paper. We use our model to develop a mechanistic understanding of how
insecticide resistance is expected to increase with the use of
insecticides in isolation, sequence and mixtures. The model flexibly
represents two independent genes coding for resistance to two
insecticides. We look principally at the ability of the insecticides to
kill susceptible mosquitoes, how much resistance counteracts this, the
proportion of mosquitoes that are exposed to insecticides and costs of
resistance.

@@

We use the model to demonstrate the evolution of resistance under
different inputs and how this fits with intuitive reasoning about the
selection pressure exerted by insecticides. We show that an insecticide
used in a mixture, relative to alone, always prompts slower evolution of
resistance, but resistance to the two insecticides may evolve more
slowly when used in sequence. We show that the ability of insecticides
to kill susceptible mosquitoes (effectiveness), has the most influence
on whether resistance to two insecticides is likely to arise faster in a
mixture or sequence.

@@

Our model makes more open the process of insecticide resistance
evolution and how it is likely to respond to insecticide use. We provide
a simple user-interface allowing further exploration
(\url{https://andysouth.shinyapps.io/MixSeqResist1}). These tools can
contribute to operational decisions in insecticide resistance
management.

@@

The WHO produced a Global Plan for Insecticide Resistance Management in
malaria vectors (GPIRM){[}1{]} which includes recommendations on
operational strategies for managing resistance including the use of
insecticide mixtures when they become available. Efforts are under way
to develop new insecticides that will be effective in the light of
existing resistance and allow additional options within insecticide
resistance management. The Innovative Vector Control Consortium (IVCC)
was set up in 2005 to develop new vector control tools and particularly
new insecticides to address insecticide resistance in malaria
transmitting mosquitoes {[}8{]}{[}9{]}. Three new insecticides are now
in development {[}9{]} and likely to be available around 2020 {[}2{]}.
It is important that decisions about how best to use the new
insecticides to delay the onset of resistance are made before
insecticides are released {[}3{]}.

Modelling studies have investigated the evolution of insecticide
resistance in insecticide mixtures including in a public health context
e.g. {[}10{]}{[}11{]}{[}12{]} but much of the work was done more than 20
years ago and there remained some confusion about the results {[}13{]}.
In an earlier paper {[}13{]} we described the technical details of a
flexible model used to investigate the relative benefits of mixtures and
sequences. Here we provide an accessible summary of the model and use
selected parameter values to describe mechanistically how the evolution
of resistance is influenced by different inputs. This mechanistic
understanding can contribute to the debate on the relative merits of
different insecticide strategies extending existing frameworks
{[}12{]}{[}4{]}{[}5{]}{[}1{]}.

Our modelling approach focuses on the change in frequency of a single
resistance gene for each insecticide. It is likely that genes conferring
resistance are present in populations even prior to exposure to novel
insecticides thus leading to the potential for selection {[}14{]}.

@@

The model outputs the change in resistance allele frequencies over time
measured in generations. We extract the number of generations at which a
resistance allele frequency of 50\% is reached and term this
`time-to-resistance'. Using a 50\% threshold is consistent with previous
modelling analyses e.g. {[}10{]}{[}15{]}. Preliminary investigations
showed that using other potential thresholds of 25 or 75\% gave very
similar results.

@@

The model is implemented in R {[}16{]}, the code is hosted on Github
{[}17{]} and we provide an on-line user interface {[}18{]} (Fig 2)
enabling the reader to change inputs and run the model themselves.

@@

When two insecticides are used we can compare the relative performance
of a mixture with sequential use. Each panel in Figures 5-9 compares
mixture and sequence for a single combination of inputs. To help
understand the mechanisms influencing this relative performance we start
with a single base scenario and demonstrate changing inputs
individually. In this base scenario we set effectiveness, exposure,
resistance restoration, and dominance at 0.5, the starting frequencies
of resistance to 0.01 and cost of resistance to 0. Resistance arises
slower for sequential than mixture for this base scenario (Fig 5A).
Resistance to both insecticides in the mixture follow the same path and
reach the threshold at the same time (as would be expected given that
they have identical input parameters). When used in sequence the curve
(dashed) for each insecticide individually is steeper than in the
mixture but, because they happen one after the other, it takes longer
for both to reach the resistance threshold (Fig 5A). Thus the
time-to-resistance for the mixture divided by the sequence is less than
1 at 0.8.

@@

The results is that a more effective insecticide increases
time-to-resistance when used in a mixture (compare the solid lines in
Fig 5A \& B). This is opposite to what happens when used in sequence
(compare the red dashed line in Fig 5A \& B) and in isolation (Fig 2A),
when more effective insecticides shortened times-to-resistance.

@@

Resistance responds to insecticide use in the model in ways that we can
explain mechanistically through the process of selection. This
mechanistic explanation can help us develop a more robust understanding
of how we would expect resistance to evolve in the field.

When single insecticide use was investigated (Figs 3-4) higher values of
inputs that increased the selective advantage of resistance all led to
more rapid evolution resistance, and higher values for cost led to
slower evolution of (Fig 4B).

@@

Using two insecticides either in sequence or in mixtures differs from
the use of a single insecticide and, where it differs can again help us
to understand the dynamics of resistance evolution. Our base scenario
(Fig 5A) shows curves for resistance to both insecticides in the mixture
that are shallower than those for sequential use. The shallower curves
for the mixture can be explained by each insecticide killing individuals
that are resistant to the other insecticide. Thus each insecticide
reduces the selection pressure for the increase in resistance to the
other and could be said to `protect' the other. However this protection
is not necessarily enough to ensure that resistance arises more slowly
in a mixture relative to a sequential strategy. Resistance can evolve
more slowly for both insecticides in sequential use, despite being
faster for each, because they occur one after the other. In Figure 5A it
takes 60 generations for resistance to reach the threshold for each
insecticide in the sequence, and 90 when they are applied together in
the mixture, so the sequence is still favoured because 2x60 is greater
than 90. Incraesing the effectiveness of on insecticide to Fig 5B
decreases the added times for the sequence to 90 and increases those for
the mixture to 115 thus changing the order and favouring the mixture.

The mechanism for this change can be sen by comparing the shape of the
curves in Figs 5A \& B. In the mixture, resistance to insecticide 1
(with the increased effectiveness) rises at a similar speed in both
figures. However, the increased effectiveness of insecticide 1 causes
resistance to insecticide 2 to increase more slowly initially in the
mixture. Once insecticide 1 reaches resistance of around 50\% at around
50 generations, the curve for insecticide 2 becomes steeper. Once the
first insecticide has become ineffective due to resistance it stops
killing individuals that are resistant to the second insecticide, loses
its protective effect and thus stops slowing the rise in resistance to
the second insecticide. A similar effect is visible when the
effectiveness of both insecticides are increased (Fig 6C). The identical
resistance curves for each insecticide in the mixture are shallower than
at the lower effectiveness (Fig 5A) because more individuals resistant
to each insecticide are being killed by the other insecticide. In this
case the `protection' given to both insecticides by the other declines
at the same rate so we don't see the change in slope seen in Fig 5B.

Fig 5B also demonstrates that when two insecticides have a different
effectiveness in a mixture, other parameters being equal, resistance
will be expected to increase faster to the more effective insecticide.
The more effective insecticide prompts both a) higher selection pressure
for it's own resistance and b) greater `protection' reducing the rise in
resistance to the other insecticide.

@@

It seems initially counter-intuitive that whereas increasing
effectiveness slows the development of resistance for mixtures,
increasing exposure speeds it up. Increasing the exposure to both
insecticides results in a decrease in time-to-resistance (from 90
generations in Fig 5A to 40 generations in Fig 5C) where increasing the
effectiveness of both insecticides by the same amount results in a
slight increase (from 90 in Fig 5A to 115 in Fig 5B). This contrasts
with the identical effects that exposure and effectiveness have on a
single insecticide (Figs 3 A,B) and on insecticides in sequence (the red
dashed lines are the same in Figs 5B \& 5C). This difference, that
exposure reduces times-to-resistance for a mixture and effectiveness
increases it, was unexpected for us.

@@

Costs of resistance favour mixtures in these analyses because the
evolution of resistance is slowed more by cost in a mixture than in a
sequence (compare the red solid lines for a mixture to the red dotted
line for a sequence in Fig 9C and D). It seems that the protective
effect of the other insecticide in a mixture interacts with the cost of
resistance to reduce selection pressure more than in a sequence.

The effects of cost of resistance become more as the frequency of
resistance increases. The resistance curves for two insecticides in a
mixture that differ only in their cost of resistance, diverge more as
the frequency of resistance increases (Fig 9C). This makes intuitive
sense, as the frequency of resistance genes increases the impact of
fitness costs associated with those genes becomes greater. Thus
differences due to cost are less noticeable at low resistance
frequencies and become more noticeable at higher resistance frequencies.
Thus if a lower resistance threshold was used cost would have less of an
effect and if a higher resistance threshold was used cost would have
more of an effect. Also note that if the costs are set even higher than
we set them here (e.g.~above 0.2) they can maintain resistance
frequencies below 1 or lead to their decline. Unfortunately, we do not
expect such high costs to occur in operational settings.

Relatedly, when there are costs of resistance, the frequency of
resistance for insecticides that are not being used is likely to
decline. Thus the frequency of resistance for the first insecticide in a
sequence, after it has been replaced, may decline to a level where it
becomes operationally useful again. As such a repeated sequential
strategy could offer benefits over mixtures not considered here. To
incorporat this one would need to use a more detailed metric than
time-to-resistance.

Our model examines solely the evolution of resistance and not the
control of mosquitoes or malaria. Different insecticide use strategies,
as well as effecting the evolution of resistance, will also influence
both the numbers of vectors and their propensity to transmit the disease
{[}19{]}. Insecticide resistance could reduce vector transmission of
disease (e.g.~by reduced lifespan of the vector) or increase
transmission (e.g.~by reduced immunity of the vector to the disease)
{[}19{]}. Our model would predict that the best strategy for limiting
the development of resistance in the short term would be to use no
insecticide. Of course, using no insecticide could have serious
consequences for mosquito and malaria abundance. The challenge is to
develop good strategies for delivering mosquito and malaria control in
the short term while sustaining their operational capacity in the longer
term by restricting the development of resistance. Others are combining
models of population genetics and population dynamics to investigate
these trade-offs {[}20{]}.

In our model we represented insecticide interaction as being additive
within the mixture. This is consistent with recent work on aphids
{[}21{]}, although that work did also indicate some synergistic effects.
Our model has the potential to incorporate either synergistic or
antagonistic effects too.

The development of insecticide resistance in operational field settings
is subject to considerable uncertainty {[}2{]}. The model we present
here is able to cover the range of that uncertainty, yet in this paper,
to develop an understanding of the key mechanisms we restrict our
explorations to a limited region of parameter space (e.g.~holding most
inputs constant at intermediate values while modifying single inputs in
isolation). Thus there is the potential that the system may behave
differently in the field than we describe here. However, the results
presented here are supported by an earlier analysis where we ran 10,000
unique scenarios varying more inputs {[}13{]}. This previous analysis
also highlighted the relative importance of insecticide effectiveness
and exposure in determining the relative performance of mixtures and
sequences. In addition we provide an on-line user interface to the model
{[}18{]} and enabling readers to investigate model behaviour in areas of
parameter space we have not shown.

@@

A subsequent model {[}22{]} lead to the following more pessimistic
conclusion about the potential of mixtures : ``As a result of incomplete
coverage and residue decay, the mortality of susceptible homozygotes is
rarely consistently high enough for pesticide mixtures to be effective''
{[}22{]}. This ``mortality of susceptible homozygotes'' is equivalent to
our insecticide effectiveness. Again this points to the importance of
effectiveness and the question of whether the effectiveness levels
needed to make mixtures better are attainable operationally. Our result
showing that decreasing exposure favours mixtures over sequences is also
consistent with figure 3 from that paper {[}22{]}.

One consistent aspect of the results presented here can seem to
contradict published advice and has been questioned when we have
presented model results. Our model predicts that lower insecticide
effectiveness or exposure will lead to slower spread of resistance in
sole use and sequences (Figs 3 \& 5, Table 2). This is consistent with
expectations that lower effectiveness and exposure would lead to lower
mortality of susceptibles and thus lower selection pressure for
resistance {[}22{]}. This can seem to contradict recommendations that
poor implementation of insecticide interventions can promote the
development of resistance. e.g.~from {[}5{]} ``Certain pest control
practices have consistently been shown to exacerbate the loss of
susceptible pest populations and the development of resistance. These
include: \ldots{} - the use of application rates that are below or above
those recommended on the label, - poor coverage of the area being
treated \ldots{}''.

Poor implementation would likely cause lower effectiveness of and
exposure to insecticides, so how can this increase the potential for
resistance when our model suggests it would be decreased ? The
explanation for this apparent difference is in the effect of dominance.
Poor application can reduce the mortality of heterozygotes (RS) thus
effectively increasing the dominance of the resistant allele and
increasing its rate of evolution {[}23{]}{[}5{]}. Our model results are
consistent with this, showing that increasing dominance of restoration
leads to faster spread of resistance (Fig 3C, 7). So, the effect of poor
implementation of an insecticide intervention on resistance could be
either positive or negative depending on the relative effects on
dominance of the resistance gene, insecticide effectiveness and
exposure. Similarly, the effects of poor application on the relative
benefits of mixtures and sequences will depend on these trade-offs
because of our result that effectivenes favours mixtures, exposure
favours sequences and higher dominance favours neither (Table 4).

\paragraph{Implications}\label{implications}

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Insecticide resistance evolution with mixtures and sequences : a model-based introduction.},
            pdfauthor={Andy South and Ian Hastings},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Insecticide resistance evolution with mixtures and sequences : a
model-based introduction.}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Andy South and Ian Hastings}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{2017-08-11}

@@

Target : Malaria journal or Parasites \& Vectors

@@

\emph{must be 350 words, currently 349}

@@

Insecticide resistance threatens effective vector control, especially
for mosquitoes and malaria. Insecticide use strategies that have been
recommended to manage resistance include mixtures, sequences and
rotations. New insecticides are being developed to address resistance
and there is an opportunity to develop use strategies that limit the
evolution of further resistance in the short term. There is limited
recent, accessible modelling work addressing the evolution of resistance
under different operational strategies. Previous work concludes that
preferred strategies will be situation specific and there appears to be
little recent consensus e.g.~on the merits of mixtures. There is an
opportunity to improve the level of mechanistic understanding within the
operational community of how insecticide resistance can be expected to
evolve in response to different strategies.

We provide a concise, accessible description of the use of a flexible
model to develop a mechanistic picture the evolution of insecticide
resistance and how it is likely to respond to potential insecticide use
strategies. The aim is to reach an audience unlikely to read a more
detailed modelling paper by assuming little knowledge of models. The
model itself, as described here, represents two independent genes coding
for resistance to two insecticides. This allows us to represent the use
of insecticides in isolation, sequence and mixtures.

@@

We use the model to demonstrate the evolution of resistance under
different inputs and how this fits with intuitive reasoning about
selection pressure. We show that using an insecticide in a mixture,
relative to alone, always prompts slower evolution of resistance to that
insecticide. However, when the resistance to both insecticides is
considered, resistance thresholds may be reached later for a sequence
relative to a mixture. Increasing the ability of insecticides to kill
susceptible mosquitoes (effectiveness), has the most influence on
favouring a mixture over a sequence.

@@

Our model provides an accessible description of the process of
insecticide resistance evolution and how it is likely to respond to
insecticide use. We provide a simple user-interface allowing further
exploration (\url{https://andysouth.shinyapps.io/MixSeqResist1}). These
tools can contribute to an improved discussion about operational
decisions in insecticide resistance management.

@@

The WHO produced a Global Plan for Insecticide Resistance Management in
malaria vectors (GPIRM){[}1{]} which includes recommendations on
operational strategies for managing resistance including the use of
insecticide mixtures when they become available. Efforts are under way
to develop new insecticides that will be effective in the light of
existing resistance and allow additional options within insecticide
resistance management. The Innovative Vector Control Consortium (IVCC)
was set up in 2005 to work with industry in the development of new
vector control tools and particularly new insecticides with new modes of
action to address insecticide resistance in malaria transmitting
mosquitoes {[}8{]}{[}9{]}. Three new insecticides are now in development
{[}9{]} and likely to be available after 2020 {[}2{]}. It is important
that decisions about how best to use the new insecticides, with a clear
mindset of delaying the evolution of resistance, are made before
insecticides are released {[}3{]}.

Modelling studies have investigated the evolution of insecticide
resistance in insecticide mixtures including in a public health context
e.g. {[}10{]}{[}11{]}{[}12{]} but much of the work was done more than 20
years ago and there remained some confusion about the results {[}13{]}.
In an earlier paper {[}13{]} we described the technical details of a
flexible model used to investigate the relative benefits of mixtures and
sequences. Here we provide an accessible summary of the model and use
selected parameter values to describe mechanistically how the evolution
of resistance can be expected to change in different scenarios. This
mechanistic understanding can contribute to the debate on the relative
merits of different insecticide strategies extending existing frameworks
{[}12{]}{[}4{]}{[}5{]}{[}1{]}.

Our modelling approach focuses on the change in frequency of a single
resistance gene for each insecticide. We assume both that genes
conferring resistance are already present in the population and that
resistance to each insecticide is coded by a single gene. These
assumptions are consistent with previous modelling work. In support of
the first assumption insecticide resistance has been termed
`pre-adaptive' meaning that resistance alleles are present in very low
numbers even prior to exposure to novel insecticides {[}14{]}{[}15{]} as
has been documented in blowflies {[}Hartley2006{]}. In support of using
a single gene representation of resistance evidence suggests that whilst
polygenic resistance is common, control failures, particularly those due
to target site resistance, are mostly due to single major genes
{[}14{]}.

@@

The model outputs the change in resistance allele frequencies over time
measured in generations. We extract the number of generations at which a
resistance allele frequency of 50\% is reached and term this
`time-to-resistance'. Using a 50\% threshold is consistent with previous
modelling analyses e.g. {[}10{]}{[}16{]}. Preliminary investigations
showed that using other potential thresholds of 25 or 75\% gave very
similar results.

@@

The model is implemented in R {[}17{]}, the code is hosted on Github
{[}18{]} and we provide an on-line user interface {[}19{]} (Fig 2)
enabling the reader to change inputs and run the model themselves.

@@

When two insecticides are used we can compare the relative performance
of concurrent use in a mixture with sequential use. Each panel in
Figures 5-9 compares mixture and sequence for a single combination of
inputs. To help understand the mechanisms influencing this relative
performance we start with a single base scenario and demonstrate
changing inputs individually. In this base scenario we set
effectiveness, exposure, resistance restoration, and dominance at 0.5,
the starting frequencies of resistance to 0.01 and cost of resistance to
0. Resistance arises slower for sequential than mixture for this base
scenario (Fig 5A). Resistance to both insecticides in the mixture follow
the same path and reach the threshold at the same time (as would be
expected given that they have identical input parameters). When used in
sequence the curve (dashed) for each insecticide individually is steeper
than in the mixture but, because they happen one after the other, it
takes longer for both to reach the resistance threshold (Fig 5A). Thus
the time-to-resistance for the mixture divided by the sequence is less
than 1 at 0.8.

@@

The result is that a more effective insecticide increases
time-to-resistance when used in a mixture (compare the solid lines in
Fig 5A \& B). This is opposite to what happens when used in sequence
(compare the red dashed line in Fig 5A \& B) and in isolation (Fig 2A),
when more effective insecticides shortened times-to-resistance.

@@

Resistance responds to insecticide use in the model in ways that we can
explain mechanistically through the process of selection. This
mechanistic explanation can help us develop a more robust understanding
of how we would expect resistance to evolve in the field. We start by
explaining evolution of resistance to a single insecticide, which is
relatively straightforward, and move on to explaining the effect of two
insecticides in sequences and mixtures.

When single insecticide use was investigated (Figs 3-4) higher values of
inputs that increased the selective advantage of resistance all led to
more rapid evolution resistance. Higher values for cost led to slower
evolution of (Fig 4B).

@@

Using two insecticides either in sequence or in mixtures differs from
the use of a single insecticide and, where it differs can again help us
to understand the dynamics of resistance evolution. Our base scenario
(Fig 5A) shows curves for resistance to both insecticides in the mixture
that are shallower than those for sequential use. The shallower curves
for the mixture can be explained by each insecticide killing individuals
that are resistant to the other insecticide. Because resistant
individuals are being killed, the selection pressure for that resistance
is lower. Thus each insecticide reduces the selection pressure for the
increase in resistance to the other and could be said to `protect' the
other. This protection means that evolution of resistance to one
insecticide is always slower when it is used in a mixture than when used
alone. This slower evolution of resistance to each insecticide in the
mixture is shown in Figs. 5-9 by the solid lines (mixture) always being
shallower than the dotted lines (sequence) for the same insecticide in
the same panel.

However whilst this protection ensures that resistance to each
insecticide evolves more slowly in a mixture it does not guarantee that
a mixture strategy is preferable to a sequence once the resistance to
both insecticides is taken into account. In a mixture selection pressure
is applied by both insecticides at the same time, whereas in a sequence
there is no selection for the insecticide not being used. Resistance can
evolve more slowly for both insecticides in sequential use, despite
being faster for each, because they occur one after the other. In Figure
5A it takes 60 generations for resistance to reach the threshold for
each insecticide in the sequence, and 90 when they are applied together
in the mixture, so the sequence is still favoured because 2x60 is
greater than 90.

This advantage of sequence over mixtures can be removed by increasing
the effectiveness of one of the insecticides. Increasing the
effectiveness of one insecticide (Fig 5B) decreases the added times for
the sequence from 120 to 90 and increases those for the mixture from 90
to 115 thus changing the order and favouring the mixture. The mechanism
for this change can be seen by comparing the shape of the curves in Figs
5A \& B. In the mixture, resistance to insecticide 1 (with the increased
effectiveness) rises at a similar speed in both figures. However, the
increased effectiveness of insecticide 1 causes resistance to
insecticide 2 to increase more slowly initially in the mixture. Once
insecticide 1 reaches resistance of around 50\% at around 50
generations, the curve for insecticide 2 becomes steeper. Once the first
insecticide reaches the resistance threshold of 50\% it kills fewer
individuals that are resistant to the second insecticide, loses its
protective effect and thus stops slowing the rise in resistance to the
second insecticide. A similar effect is visible when the effectiveness
of both insecticides are increased (Fig 6C). The identical resistance
curves for each insecticide in the mixture are shallower than at the
lower effectiveness (Fig 5A) because more individuals resistant to each
insecticide are being killed by the other insecticide. In this case the
`protection' given to both insecticides by the other declines at the
same rate so we don't see the change in slope seen in Fig 5B. In both
cases increasing the effectiveness of insecticides increases their
protective effect when used in mixtures and favours mixtures over
sequences.

Fig 5B also demonstrates that when two insecticides have a different
effectiveness in a mixture, other parameters being equal, resistance
will be expected to increase faster to the more effective insecticide.
The more effective insecticide prompts both a) higher selection pressure
for it's own resistance and b) greater `protection' reducing the rise in
resistance to the other insecticide. The more effective insecticide has
a faster evolution of resistance but this is not caused by the presence
of the less effective one.

@@

It seems initially counter-intuitive that whereas increasing
effectiveness slows the development of resistance for mixtures,
increasing exposure speeds it up. Increasing the exposure to both
insecticides results in a decrease in time-to-resistance (from 90
generations in Fig 5A to 40 generations in Fig 5C) where increasing the
effectiveness of both insecticides by the same amount results in a
slight increase (from 90 in Fig 5A to 115 in Fig 5B). This contrasts
with the identical effects that exposure and effectiveness have on a
single insecticide (Figs 3 A,B) and on insecticides in sequence (the red
dashed lines are the same in Figs 5B \& 5C). This difference, that
exposure reduces times-to-resistance for a mixture and effectiveness
increases it, was unexpected for us and we are keen to explore it in
more detail.

@@

Fitness costs of resistance have been widely reported(review in
{[}20{]}) but it has been suggested that few of these are directly
relevant to control failure in the field {[}{\textbf{???}}{]}. For this
reason most of our analyses (with the exception of Fig 9) do not include
costs of resistance. It has been suggested {[}{\textbf{???}}{]} that
strategies using more than one insecticide rely on the assumption of
costs for resistance to one insecticide to decline when it is not being
used. Our model allows us to compare strategies with or without the
assumption of costs. Without costs resistance levels do stay constant
when an insecticide is not in use.

Costs of resistance favour mixtures in these analyses because the
evolution of resistance is slowed more by cost in a mixture than in a
sequence (compare the red solid lines for a mixture to the red dotted
line for a sequence in Fig 9C and D). It makes sense that the protective
effect of the other insecticide in a mixture combines with the cost of
resistance to reduce selection pressure more than in a sequence.

The effects of cost of resistance become more as the frequency of
resistance increases. The resistance curves for two insecticides in a
mixture that differ only in their cost of resistance, diverge more as
the frequency of resistance increases (Fig 9C). This makes intuitive
sense too, as the frequency of resistance genes increases the impact of
fitness costs associated with those genes becomes greater. Thus
differences due to cost are less noticeable at low resistance
frequencies and become more noticeable at higher resistance frequencies.
Thus if a lower resistance threshold was used cost would have less of an
effect and if a higher resistance threshold was used cost would have
more of an effect. Also note that if the costs are set even higher than
we set them here (e.g.~above 0.2) they can maintain resistance
frequencies below 1 or lead to their decline. However, we do not expect
such high costs to occur in operational settings.

Relatedly, when there are costs of resistance, the frequency of
resistance for insecticides in a sequence that are not being used is
likely to decline. Thus the frequency of resistance for the first
insecticide in a sequence, after it has been replaced, may decline to a
level where it becomes operationally useful again. As such a repeated
sequential strategy could offer benefits over mixtures not considered
here. To quantify these benefits one would need to use a more detailed
metric than the time-to-resistance we have used. We quantified the
number of generations until resistance thresholds are first reached.
Other measures for quantifying resistance over time can be imagined.

@@

A subsequent model {[}21{]} lead to the following more pessimistic
conclusion about the potential of mixtures : ``As a result of incomplete
coverage and residue decay, the mortality of susceptible homozygotes is
rarely consistently high enough for pesticide mixtures to be effective''
{[}21{]}. This ``mortality of susceptible homozygotes'' is equivalent to
our insecticide effectiveness. Again this points to the importance of
effectiveness and the question of whether the effectiveness levels
needed to make mixtures better are attainable operationally. Our result
showing that decreasing exposure favours mixtures over sequences is also
consistent with figure 3 from that paper {[}21{]}.

One aspect of the results presented here can seem to contradict
published advice and has been questioned when we have presented model
results. Our model predicts that lower insecticide effectiveness or
exposure will lead to slower spread of resistance in sole use and
sequences (Figs 3 \& 5, Table 2). This is consistent with expectations
that lower effectiveness and exposure would lead to lower mortality of
susceptibles and thus lower selection pressure for resistance {[}21{]}.
This can seem to contradict recommendations that poor implementation of
insecticide interventions can promote the development of resistance.
e.g.~from {[}5{]} ``Certain pest control practices have consistently
been shown to exacerbate the loss of susceptible pest populations and
the development of resistance. These include: \ldots{} - the use of
application rates that are below or above those recommended on the
label, - poor coverage of the area being treated \ldots{}''.

Poor implementation would likely cause lower effectiveness of and
exposure to insecticides, so how can this increase the potential for
resistance when our model suggests it would be decreased ? The
explanation for this apparent difference is in the effect of dominance.
Poor application can reduce the mortality of heterozygotes (RS) thus
effectively increasing the dominance of the resistant allele and
increasing its rate of evolution {[}22{]} {[}5{]}. Our model results are
consistent with this, showing that increasing dominance of restoration
leads to faster spread of resistance (Fig 3C, 7). So, the effect of poor
implementation of an insecticide intervention on resistance could be
either positive or negative depending on the relative effects on
dominance of the resistance gene, insecticide effectiveness and
exposure. Similarly, the effects of poor application on the relative
benefits of mixtures and sequences will depend on these trade-offs
because of our result that effectiveness favours mixtures, exposure
favours sequences and higher dominance favours neither (Table 4). More
research into how the mortality of different genotypes (susceptible,
resistant and heterozygous) is affected by poor application rates or
insecticide degradation over time would help make this clearer.

Experience from agriculture where single interventions have been
associated with a rapid development of resistance have led for calls for
a more Integrated Vector Management composed of a series of partially
effective tools {[}23{]}. It is suggested that such an integrated
approach is more sustainable and `evolution-proof' {[}23{]}. A similar
combining of interventions has been advocated to cover mosquitoes
exhibiting different behaviours {[}24{]}. Mosquitoes with different
behaviours e.g.~feeding outdoors and/or on animals are likely to result
in lower exposures to insecticides in nets or sprayed on walls. This
lower exposure could favour mixtures over sequences as we have
indicated. In both cases our modelling approach could be used to
investigate implications for the development of resistance.
